Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2024-08-19 manuscript/manuscript.Rmd
finished another round of proofreading.

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract:

date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: jhe.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf. Long Core Tools [@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010).
By comparing variation in overall artifact shape (or “Bauplan” [@lycettQuestionsSurroundingAcheulean2008]) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.

@@

Typologically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\>10cm) and shape (elongation and flattening) [e.g., @isaac1977].
This consistent production of large, flat, and elongated cores in the Achuelean has long been thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018].
Unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small "protobifaces" that occur in low frequencies in Oldowan assemblages [@isaac1997].
However, the Early Acheulean is clearly distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts.
Although studies of handaxe variation often focus on shape rather than size, this shift is an important aspect of artifact design with relevance to both production and function.

Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or simply by using larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018].
Both may involve similar flaking "strategies" (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes.
This increases the perceptual motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009].
These new challenges would have increased raw material procurement [@shea2010] and learning costs [@pargeter2019] as well as the risk of serious injury [@gala2023] associated with tool production.
This strongly implies intentional pursuit of offsetting functional benefits related to size increase.
These likely included tool ergonomics and performance [@key2017] as well as flake generation, resharpening, and reuse potential [@shea2010].
Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns [@linaresmatás2021; @rogers1994].

The degree of intentional design reflected in the shape of Early Acheulean LCTs is more difficult to determine.
For example, LFB production using a simple "least effort" bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982] whether or not this is an intentional design target.
Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes even in the absence of explicit design targets.
On the other hand, it has been argued that the shape of Early Acheulean LFBs was intentionally predetermined using core preparation techniques [@delatorre2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @duke2021; @lepre2011; @semaw2009].
To date, however, the identification of Early Acheulean shaping has generally relied on qualitative assessment by lithic analysts.
Such assessment may in fact be reliable, but is subject to concerns about potential selectivity, bias, and/or overinterpretation [@davidson2002; @moore2016].
Notably, a 3-dimensional morphometric (3DGM) study by Presnyakova, et al. [-@presnyakova2018] concluded that LCT shape variation in the Okote Member (\~1.4 mya) at Koobi Fora was largely driven by reduction intensity rather than different knapping strategies.
However, this study did not directly address the presence/absence of design targets constraining the observed range of variation.

In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as an alternative to intentional form imposition [@mcpherron2000].
Across heavily-worked and relatively standardized LCT assemblages [e.g., @shipton2020], a *lack* of association between morphology and reduction intensity has be used as an argument-by-elimination for the presence of imposed morphological norms or "mental templates" [@garcía-medrano2019; @shipton2015; @shipton2023].
However, in the less heavily-worked and more heterogeneous assemblages typical of the early Acheulean [@kuhn2020], it is equally plausible that increasing reduction intensity would reflect degree of primary reduction rather than subsequent resharpening [@archer2010].
In this case, reduction intensity effects on morphology would have the opposite interpretation: more reduction should result in closer approximation of a desired form if such were present.
For example, Beyene, et al. [-@beyene2013] found that increasing flake scar counts were associated with increasing handaxe refinement through time at Konso, Ethiopia, which may reflect a more general trend in the African Acheulean [@shipton2018].

Interpretive approaches address this quandary by "reading" the organization of scars on individual pieces to infer intent, but an adequate method to objectively quantify these insights has yet to be developed.
Current measures of reduction intensity, such as the scar density index (SDI) [@clarkson2013], are designed to estimate total mass removed from a core and are reasonably effective [@lombao2023].
However, mass removal was not the objective of Paleolithic flaking.
Indeed, knapping efficiency is usually conceived as generating an outcome while *minimizing* required mass removal.
This is true whether the desired outcome is a useful flake, a rejuvenated edge, or a particular core morphology.
In simple flake production, mass removed is probably a good reflection of the completeness of exploitation ("exhaustion") of cores and may have implications for required skill [@toth1982; @pargeter2023] as well as raw material economy [@reeves2021; @shick1987].
However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized, and might even interfere with function [@keyHandaxeShapeVariation2019].
Without further information, relationships between artifact shape and reduction intensity are thus open to conflicting interpretations as evidence of intentional design or its absence.

Li, et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
However, they also explain that "flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p. 6).
We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
(**Figure** \@ref(fig:FAISDI)) In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.

```{r FAISDI, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A conceptual model of inferring intention from artifact morphology."}
knitr::include_graphics(here::here("figure/FAI_SDI.png"))
```

@@

Three-dimensional scanning and geometric morphometric (3DGM) methods are increasingly common in the study of LCT form and reduction intensity [@shipton2015; @caruana2020; @presnyakova2018; @li2021; @archer2010; @li2015; @lycett2006].
These methods provide high-resolution, coordinate-based descriptions of artifact form including detailed information about whole object geometric relations that is not captured by conventional linear measures [@shott2010].
This includes measures of surface area used to compute both SDI and FAI measures [@clarkson2013; @li2015].
At the time of writing, however, 3D scans are available for only a small number of Gona artifacts, including 33 of the Oldowan and Acheulean flaked pieces used in this study (**Figure** \@ref(fig:fig1)).
Despite continuing improvements, 3DGM methods still impose additional costs in terms of data collection and processing time as well as required equipment, software, and training.
Importantly, 3DGM methods cannot be applied to pre-existing photographic and metric data sets [e.g., @marshall], including available data from Gona.
For this reason, and to better understand the relative costs and benefits of 3DGM more generally, we sought to test the degree to which conventional measurements can approximate 3DGM methods and produce reliable results by directly comparing our conventional measures with 3DGM analysis of the 33 available scans.

```{r fig1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D Artifact scans from Gona used in this study."}
knitr::include_graphics(here::here("figure/Fig1.png"))
```

For our study, we are specifically concerned with the accurate description of morphological variation and estimation of artifact surface and flaked areas.
With respect to morphology, we were encouraged by the fact that aspects of form variation relevant to our research question (i.e., core elongation, flattening, and pointedness) are relatively simple to describe using sparse data.
3DGM studies of LCTs collect vast amounts of shape data but may discard upward of 50% of the observed variation in order to focus on two or three interpretable principal components.
Across studies, these PCs consistently correspond to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower-resolution spatial data [@garcía-medrano2019; @archer2010; @lycett2006] and studies employing linear measures rather than spatial coordinates [@pargeter2019; @crompton1993].
There is less evidence that conventional methods can accurately estimate artifact surface and flaked areas.
Clarkson [-@clarkson2013] advocated the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions).
However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive r^2^ = 0.944 and no effect of variation in core shape.
Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than absolute artifact size, such systematic overestimation may not be problematic.
Similar concerns apply to the estimation of flaked area.
Traditionally, such estimates have been done "by eye" as a percentage of the total artifact surface [e.g., @dibble2005].
Such estimations have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010].
The accuracy of by-eye estimation has yet to be systematically studied in Early Stone Age cores like those from Gona.

@@

Early Acheulean sites in the Gona Project area (**Figure** \@ref(fig:fig2)) are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7 to 1.2 mya [@semaw2018].
The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12, and BSN-17, all estimated to ca.
1.7 -- 1.4 mya by stratigraphic position in the Gona sequence [@quade2008; @semaw2020].
The Busidima Formation accumulated through fluvial deposition by the ancestral Awash River (Type I context) and its smaller tributary channels (Type II context) [@quade2004; @quade2008].
Oldowan sites at Gona all occur in Type I sediments, indicating channel bank/margin (OGS-7) or proximal floodplain (EG-10, EG-12) contexts close to the large, hetero-lithic clasts available from point bars in the axial river channel [@quade2004; @stout2005].
Acheulean sites continue to occur in Type I contexts (BSN-17, DAN-5) but are also found in Type II sediments (OGS-5, OGS-12,) reflecting increased utilization of large perennial tributaries to the ancestral Awash River [@quade2008].
Clasts locally available in these tributaries were relatively small, implying that the large flakes and cobbles used to produce Acheulean artifacts were initially sourced from the axial river.
A similar pattern of habitat diversification and increasing lithic transport distances has been described at other sites and may be typical of the early Acheulean [@rogers1994; @linaresmatás2021; @hay1976].
As with other early (i.e. \>1.0 mya [@stoutStoneToolmakingEvolution2011; @presnyakova2018]) Acheulean assemblages, the Gona collections include numerous "crudely made" handaxes and picks on large flake blanks and cobbles, as well as large (\> 10cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores (see Figure 1af), and smaller debitage [@semaw2018; @semaw2020].

<!-- ```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."} -->

<!-- knitr::include_graphics(here::here("figure/site map.png")) -->

<!-- ``` -->

# Materials and Methods

## Materials

### Archaeological Sample

Artifact collections analyzed here include *in situ* pieces excavated from intact stratigraphic contexts and surface pieces systematically collected from the sediments eroding from these layers.
Surface pieces are included because the current technological analysis does not require more precise spatial association, stratigraphic, and chronological control.
Our sample comprises the total collection of flaked pieces [@isaac1997] and large (\>10 cm) detached pieces from each site, regardless of typo-technological interpretation.

## Methods

### Artifact Classification

>>>>

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract:
|
  ```{=latex}
  The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities. However, the nature and extent of explicit design documented by these forms remains open to debate. To address this issue, we analyzed the complete collection of early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared these with all of the flaked pieces from two published Oldowan (>2.5 Ma) sites at Gona. By comparing shape variation to measures of flaking intensity and coverage, we sought to identify technological patterns indicative of intent. Results provide little support for the presence of discrete tool types or imposed morphological norms in our sample. Instead, we observe systematic patterns of raw material selection and core surface modification aimed at the production and maintenance of useful cutting edges on relatively large supports (cobbles and large flake blanks). This is consistent with prior characterizations of Acheulean tool form as arising from functional and ergonomic design imperatives for large hand-held cutting tools. Although the generalizability of these results to other sites remains to be seen, we propose that distinctive early Acheulean artifact forms may represent secondary accommodations to the primary goal of increasing tool size to meet the novel demands (e.g., extended use-life, enhanced transportability, utility for heavy-duty cutting) of a more general shift in hominin behavioral ecology at this time. Our results thus provide support for the presence of Acheulean design, not as the concrete realization of a ‘holographic model’ in the mind of the maker, but rather in the broader sense of deliberate technological choices made in view of behavioral goals and material constraints.

  \textbf{Keywords:} Lithic technology; Early Stone Age; Handaxe; Large Cutting Tool; Pick; Shaping
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: jhe.csl
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” [LCTs; cf. Long Core Tools; @shea2013] in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969]. However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial. Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6]. Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design. Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020]. Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016]. To examine these questions, we studied the complete collection of Early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\> 2.5 Ma) sites at Gona [OGS7 and EG10; @semaw2000; @semaw2003; @stoutTechnologicalVariationEarliest2010]. By comparing variation in overall artifact shape (or “Bauplan” [@lycettQuestionsSurroundingAcheulean2008]) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent. The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.

@@

Classically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\> 10 cm) and basic aspects of overall artifact shape including elongation and flattening [@isaac1977] and the creation of a point [@kleindienst1962] . The consistent production of large, flat, elongated, and pointed forms in the Acheulean is generally thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018]. Within this broad frame, numerous studies have examined the variation in overall LCT shape captured by various combinations of length, breadth, and thickness measures in an attempt to disentangle cultural, functional, raw material, allometric, and resharpening influences [e.g., @roe1969; @crompton1993; @mcpherron2000]. It has been demonstrated that unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small “protobifaces” that occur in low frequencies in Oldowan assemblages [@isaac1997]. However, the early Acheulean is also distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts. Although studies of handaxe variation often focus on shape rather than size, this size increase is an important aspect of artifact design with relevance to both production and function.

Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or by selecting larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018]. Both approaches may involve similar flaking “strategies” (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes. This increases the perceptual-motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009]. These new challenges would have increased raw material procurement [@shea2010] and learning [@pargeter2019] costs, as well as the risk of serious injury [@gala2023]. These costs strongly imply intentional pursuit of offsetting functional benefits related to size increase. Such benefits likely included tool ergonomics and performance [@key2017; @keyReassessingProductionHandaxes2017], as well as flake generation, resharpening, and reuse potential [@shea2010]. Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns[@linaresmatás2021; @rogers1994; @hay1976]. Importantly, such functional and ecological interpretation speaks primarily to motivation (goals), and need not imply a lack of hominin technological understanding and intention.

The degree of intentional design reflected in the overall shape of Early Acheulean LCTs is more difficult to determine. For example, LFB production using a simple “least effort” bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982; @leader2018] whether or not this is an intentional design target. Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes, even in the absence of normative shape targets. Nevertheless, it has been argued that the shape of Early Acheulean LFBs was predetermined using intentional core preparation techniques [@delatorre2018; @leader2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018; @presnyakova2018].

To date, evidence of early Acheulean shaping has generally come from qualitative assessments by lithic analysts. We do not dispute these assessments, but note that they are subject to reasonable concerns about possible selectivity, bias and/or overinterpretation [e.g., @davidson2002; @moore2016]. The attribution of overall artifact shape imposition would be more compelling if supported by multiple lines of qualitative and quantitative evidence [i.e., a consilience of inductions; @whewell1837]. This is particularly true as qualitative and quantitative approaches often have complementary strengths and weaknesses. Currently, however, quantitative methods bearing on this question are underdeveloped and evidence is mixed. A three-dimensional geometric morphometric (3DGM) study by Presnyakova et al. [-@presnyakova2018] concluded that LCT shape variation in the Okote Member (ca. 1.4 Ma) at Koobi Fora was largely driven by reduction intensity rather than different knapping strategies. However, this study did not directly address the presence/absence of design targets constraining the observed range of variation. Shipton et al. [-@shipton2019] found that there was a correlation between LCT symmetry and reduction intensity (scar density index) in the lightly worked cleavers from Olorgesailie CL1-1 and conclude that this, along with several examples of flake scar patterning, provides evidence of the intentional imposition of symmetry. However, they remain agnostic as to whether this was motivated by functional, aesthetic, or social goals, leaving open the possibility that overall symmetry was a byproduct of more localized morpho-functional objectives. The purpose of the current study is to add to these various lines of evidence by developing and applying novel quantitative methods for assessing the imposition of overall artifact form in the early Acheulean at Gona.

In later Acheulean contexts, reduction intensity effects are commonly equated with resharpening and seen as supporting an alternative “byproduct” explanation of artifact shape regularity, rather than intentional form imposition [@mcpherron2000]. Across heavily worked and relatively standardized LCT assemblages [e.g., @shipton2020], a lack of association between morphology and reduction intensity has thus been used as an argument-by-elimination for the presence of imposed morphological norms or “mental templates” [@garcía-medrano2019; @shipton2015; @shipton2023]. However, in the less heavily worked and more heterogenous assemblages typical of the early Acheulean [@kuhn2020], reduction intensity is more plausibly interpreted as an indicator of the extent of primary reduction rather than subsequent resharpening [@archer2010; @shipton2019]. In this case associations between reduction intensity and shape would indicate intended effects of primary reduction, rather than byproducts of subsequent rounds of resharpening. For example, Beyene et al. [-@beyene2013] found that increasing flake scar counts were associated with increasing handaxe refinement through time at Konso, Ethiopia, which may reflect a more general trend in the African Acheulean [@shipton2018]. Disentangling such alternative explanations is a major challenge for quantitative approaches to the question of shape imposition.

Interpretive approaches address this quandary by “reading” the organization of scars on individual pieces to infer intent [@pelegrin2005], but an adequate method to objectively quantify these insights has yet to be developed. Lin et al. [-@lin2024] recently developed a method for quantifying flake scar orientations patterns, however this method still requires reading the direction of individual scars and it is not yet clear how results might be related to the question of shape imposition. The WEAP method of Garcia-Medrano et al. [-@garcia-medranoWEAPMethodNew2020] similarly provides a useful framework for integrating qualitative/interpretive technological attributes with quantitative shape metrics but does not address the reliability of individual attributes. Current measures of reduction intensity, such as the scar density index [SDI; @clarkson2013], are designed to estimate total mass removed from a core and have been shown to be reasonably effective [@lombao2023]. However, mass removal was not the objective of Paleolithic flaking. Indeed, knapping efficiency is usually conceived as generating an outcome while minimizing required mass removal and energy expenditure [@rezek2018]. This is true whether the desired outcome is a useful flake, a rejuvenated edge, or a particular core morphology. In simple flake production, mass removed is probably a good reflection of the completeness of exploitation (“exhaustion”) of cores and may have implications for required skill [@toth1982; @pargeter2023], as well as raw material economy [@reeves2021; @shick1987]. However, in core shaping and resharpening, mass removal would typically represent an energetic and raw material cost to be minimized and might even interfere with function [@keyHandaxeShapeVariation2019]. Without further information, relationships between artifact shape and reduction intensity are thus open to conflicting interpretations as evidence of intentional design or its absence.

Li et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by a moderate correlation ($r = 0.424$) with SDI. However, they also explain that “flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage.” [@li2015: 6]. We suggest that FAI actually captures the spatial extent of modifications to the surface of a core, rather than the volume of material removed. It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations (**Fig.** \@ref(fig:fig1)). For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on “least-effort” shape imposition through large and/or strategically placed flake removals. On the other hand, an association between SDI and core form with no effect of FAI would be consistent with reduction (shaping and/or resharpening) focusing on spatially restricted areas (e.g., non-invasive retouch) on the core. In both cases we would also expect a breakdown in the expected association between FAI and SDI. For simple debitage with no systematic morphological constraints we would expect a lack of any effect of FAI or SDI on core shape. Conversely, strong correlation of core shape with both FAI and SDI would indicate a highly “designed” form achieved through extensive surface modification and volumetric reduction. In the current study we employed these theoretical expectations to evaluate evidence of the presence and nature of intentional shaping in the early Acheulean of Gona.

```{r fig1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Technological interpretations of the relationship between artifact shape, flaked area, and scar density. Abbreviations: SDI: corrected scar density index; FAI: Flaked area index."}
knitr::include_graphics(here::here("figure/Fig.1.png"))
```

@@

For our study, we are specifically concerned with the accurate description of morphological variation and estimation of artifact surface and flaked areas. Historically, debates over imposed form in the Acheulean have focused on basic aspects of overall artifact shape, such as elongation, flattening, and pointedness that are relatively easy to capture using sparse data. For example, McNabb [-@mcnabb2022] recently concluded that the influential “Roe method” using 7 linear measurements is sufficient to “fairly represent the basic *bauplan* of LCTs.” Typological groupings derived in this fashion have similarly been validated [@shipton2020] by analyses based on the much more detailed morphological information provided by three-dimensional geometric morphometric (3DGM) methods. Across 3DGM studies, the first two or three principle components (PCs) typically account for about 50% of variation and consistently correspond to basic features like elongation, relative thickness, pointedness, and position of maximum thickness that also emerge from lower resolution spatial data [@lycett2006; @archer2010; @garcía-medrano2019] and studies employing linear measures rather than spatial coordinates [@crompton1993; @pargeter2019]. We are thus confident that the seven linear measures used in our study effectively capture relevant aspects of shape variation for our research question. To further confirm this, we compared shape PCs derived from our linear measures with the results of a 3DGM analysis of all scanned artifacts in our sample ($n = 31$; **Fig.** \@ref(fig:fig2)).

```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D Artifact scans from Gona used in this study.Artifact maximum dimension is vertical, and three views shown 90 degree rotation. The pieces, catalog number and typological classifications are: EG10: 1125 M1 FC (a), 1171 M1 FC (b), 1215 M1 FC(c), 1223 M1 FC (d). OGS7: 334 M1 FC (e), 335 M1 FC (f), 437 M1 FF (g), 485 M1 FC (h), 486 M1 FIB (i), 2016 “M1 FF “protopick” (j), 523 M1 FC (k), 2006 M1 FF (l), 2066 M1 FC (m), 2081 M1 FIB (n), 2092 M1 FC (o). OGS12: 16 M2 FC “handaxe” (p), 63 M2 FF “handaxe” (q), 226 M2 FF “handaxe” (r), 13 M2 FF “pick” (s), 74 M2 FC “pick” (t), 102 M2 FF “pick” (u), 2012 M2 FF “pick” (v), 115 M2 FF “knife” (w), 110 M2 FF “knife” (x), 113 M2 FF “knife” (y). BSN17: 35 M2 FC “handaxe” (z), 21 M2 FC “pick”  (aa), 24 M2 FC “pick” (ab), 34 M2 FC “pick” (ac), 29 M2 FC “broken pick” (ad), 231 M1 FC “core” (ae) (M1: Mode 1, M2: Mode 2, FC: Flaked Cobble, FF: Flaked Flake, FIB: Flaked Indeterminate Base)"}
knitr::include_graphics(here::here("figure/Fig.2.png"))
```

There is less evidence that conventional methods can accurately estimate artifact surface and flaked areas. Clarkson [-@clarkson2013] advocated the use of 3D surface area measures as more accurate than estimation from linear measures (e.g., surface area of a rectangular prism defined by artifact dimensions). However, he also found that the error introduced by the linear approach was a highly systematic, isometric overestimation of surface area and that results correlated with direct 3D measures with an impressive $r^2 = 0.944$ and no effect of variation in core shape. Insofar as it is variation in the relationship between surface area and flaking intensity that is of interest, rather than absolute artifact size, such systematic overestimation may not be problematic. Similar concerns apply to the estimation of flaked area. Traditionally, such estimates have been done “by eye” as a percentage of the total artifact surface [e.g., @dibble2005]. Such estimations have been found to be reasonably accurate when compared to 3D methods, but with the potential for substantial error on individual artifacts [@lin2010]. The accuracy of visual estimation has yet to be systematically studied in Early Stone Age cores like those from Gona. We addressed this question by comparing SDI and FAI estimates derived from linear measures and “by eye” estimation with the results of 3DGM analysis.

@@

Early Acheulean sites in the Gona Project area are distributed over a wide area within the Dana Aoule North (DAN), Ounda Gona South (OGS), and Busidima North (BSN) drainages in the Busidima Formation [@quade2004] and range in age from approximately 1.7–1.2 Ma [@semaw2018]. The specific sites included in the current analysis are DAN-5, OGS-5, OGS-12 (all ca. 1.6–1.5 Ma), and BSN-17 (1.3 Ma), with ages estimated by stratigraphic position in the Gona sequence [@quade2008; @semaw2020]. The Busidima Formation accumulated through fluvial deposition by the ancestral Awash River (Type I context) and its smaller tributary channels [Type II context; @quade2004; @quade2008]. Oldowan sites at Gona all occur in Type I sediments, indicating channel bank/margin (OGS-7) or proximal floodplain (EG-10, EG-12) contexts close to the large, hetero-lithic clasts available from point bars in the axial river channel [@quade2004; @stout2005]. Acheulean sites continue to occur in Type I contexts (BSN-17, DAN-5) but are also found in Type II sediments (OGS-5, OGS-12) reflecting increased utilization of large perennial tributaries to the ancestral Awash River [@quade2008]. Clasts locally available in these tributaries were relatively small, implying that the large flakes and cobbles used to produce Acheulean artifacts were initially sourced from the axial river. A similar pattern of habitat diversification and increasing lithic transport distances has been described at other sites and may be typical of the early Acheulean [@rogers1994; @linaresmatás2021; @hay1976]. As with other early [i.e., \> 1.0 mya; @stoutStoneToolmakingEvolution2011; @presnyakova2018] Acheulean assemblages, the Gona collections include numerous “crudely made” handaxes and picks on large flake blanks and cobbles, as well as large (\> 10 cm) unmodified flakes, flaked pieces interpreted typo-technologically as Mode 1 cores, and smaller debitage [@semaw2018; @semaw2020].

# Materials and methods

## Archaeological sample

Artifact collections analyzed here include in situ pieces excavated from intact stratigraphic contexts and surface pieces systematically collected from the sediments eroding from these layers. Surface pieces are included because the current technological analysis does not require more precise spatial association, stratigraphic, and chronological control. Our sample comprises the total collection of flaked pieces [@isaac1997] from each site, regardless of size, as well as any large (\> 10 cm) unmodified detached pieces (total $n = 226$). As summarized in Table 1, our analytic categories thus include Flaked Cobbles, Flaked Flakes, Flaked Pieces on Indeterminate bases, and Unmodified Large Flakes. Note that the Flaked Piece categories are not size dependent, and a number of small Flaked Flakes from Oldowan sites are thus included. Our intention is to avoid a priori assumptions about artifact typology, technology, or function. All artifact collections from Gona are housed at the National Museum of Ethiopia.

## Artifact classification and comparison
